An Algorithmic Introduction to Lagrangian Coherent Structures

### 2013-11-13 main.tex
Update Bickley jet section

The Bickley jet models a zonal jet flanked above and below by
stationary vortices. This is an idealised model of geophysical flows
such as the polar night jet perturbed by a Rossby
wave\parencite{beron-vera10:_invar_lagran,haller12:_geodes_theor_trans_barrier_two_dimen_flows}.

The model is defined by two stream functions, $\psi(x,y,t) = \psi_0(x,y) + \psi_1(x,y,t)$, where:
\begin{gather*}
\psi_0(x,y) = c_3 y - U L_y \tanh\frac{y}{L_y} + \epsilon_3 U L_y \mathrm{sech}^2\frac{y}{L_y} \cos k_3 x,\\
\psi_1(x,y,t) = U L_y \mathrm{sech}^2\frac{y}{L_y} \mathrm{Re}\left[ \sum_{n=1}^2 \epsilon_n f_n(t) e^{i k_n x}\right].
\end{gather*}
The forcing function is a solution of the Duffing oscillators with parameters in the chaotic regime, specifically
\begin{gather*}
\frac{d \phi_1}{dt} = \phi_2,\\
\frac{d \phi_1}{dt} = -0.1 \phi_2 - \phi_1^3 + 11 \cos(t),\\
f_{1,2}(t) = 2.625 \times 10^{-2} \phi_1(t/6.238 \times 10^5)
\end{gather*}
The velocity is given by:
\[
\boldsymbol{v}(x,y,t) = (-\partial_y \psi, \partial_x \psi).
\]
The parameter values used are: $U = 62.66$, $c_2 = 0.205 U$, $c_3 = 0.461 U$, $L_y = 1.77 \times 10^6$, $\epsilon_1 = 0.0075$, $\epsilon_2 = 0.04$, $\epsilon_3 = 0.3$, $L_x = 6.371 \times 10^6 \pi$, $k_n = 2 n \pi/L_x$, $\sigma_1 = 0.5 k_2 (c_2 - c_3)$, $\sigma_2 = 2 \sigma_1$. The domain of the flow is set to $x \in [0,L_x]$, $y \in [-4 \times 10^6, 4 \times 10^6]$ and the timespan is set to $4 L_x/U$.

@@

To calculate the Cauchy-Green strain tensor and its eigenvalues and eigenvectors, the main grid resolution is set to $500 \times 200$ and the auxiliary grid spacing is set to one percent of the main grid spacing (i.e. 400.) The main grid is used to calculate eigenvalues and flow incompressibility is imposed. Periodic boundary conditions are imposed in the horizontal direction.

To obtain shear LCS Poincare sections are specified. The maximum orbit length is set to twice the circumference circles whose radius would be the Poincare sections.

To obtain hyperbolic LCS, the local maximisation distance is set to eight times the main grid spacing and the maximum strainline length is set to $1 \times 10^8$.

>>>>

The Bickley jet models a zonal jet flanked above and below by
stationary vortices. This is an idealised model of geophysical flows
such as the polar night jet perturbed by a Rossby
wave\parencite{haller12:_geodes_theor_trans_barrier_two_dimen_flows,beron-vera10:_invar_lagran}.

The velocity is given by $\boldsymbol{v}(x,y,t) = (-\partial_y \psi, \partial_x \psi)$ and
\begin{gather*}
\psi(x,y,t) = \psi_0(x,y) + \psi_1(x,y,t),\\
\psi_0(x,y) = c_3 y - U L_y \tanh\frac{y}{L_y} + \epsilon_3 U L_y \mathrm{sech}^2\frac{y}{L_y} \cos k_3 x,\\
\psi_1(x,y,t) = U L_y \mathrm{sech}^2\frac{y}{L_y} \mathrm{Re}\left[ \sum_{n=1}^2 \epsilon_n f_n(t) e^{i k_n x}\right].
\end{gather*}
Periodic boundary conditions are imposed in the $x$-direction.The forcing function we choose is a chaotic solution of the Duffing oscillators with parameters in the chaotic regime, specifically
\begin{gather*}
\frac{d \phi_1}{dt} = \phi_2,\\
\frac{d \phi_1}{dt} = -0.1 \phi_2 - \phi_1^3 + 11 \cos(t),\\
f_{1,2}(t) = 2.625 \times 10^{-2} \phi_1(t/6.238 \times 10^5)
\end{gather*}
The parameter values used are: $U = 62.66$, $c_2 = 0.205 U$, $c_3 = 0.461 U$, $L_y = 1.77 \times 10^6$, $\epsilon_1 = 0.0075$, $\epsilon_2 = 0.04$, $\epsilon_3 = 0.3$, $L_x = 6.371 \times 10^6 \pi$, $k_n = 2 n \pi/L_x$, $\sigma_1 = 0.5 k_2 (c_2 - c_3)$, $\sigma_2 = 2 \sigma_1$. The domain of the flow is set to $x \in [0,L_x]$, $y \in [-2.25 L_y, 2.25 L_y]$ and the timespan is $t \in [0,4 L_x/U]$. We skip listing the file that defines the velocity equations; in regards to input and output variables, it is similar to the double gyre function (Listing~\ref{l:double gyre derivative}.) The LCS Tool script that specifies the parameters we are using is given in Listing~\ref{l:Bickley jet input}. Note the use of the \texttt{periodicBc} variable to specify periodic boundary conditions.

\begin{lstlisting}[caption={Bickley jet input file: \texttt{demo/bickley\_jet/hyperbolic\_shear\_lcs.m}.},label=l:Bickley jet input]
%% Input parameters
u = 62.66;
lengthX = pi*earthRadius;
lengthY = 1.77e6;
epsilon = [.075,.4,.3];
domain = [0,lengthX;[-1,1]*2.25*lengthY];

% Make x and y grid spacing as equal as possible
resolutionX = 500;
gridSpace = diff(domain(1,:))/(double(resolutionX)-1);
resolutionY = round(diff(domain(2,:))/gridSpace);
resolution = [resolutionX,resolutionY];

timespan = [0,4*lengthX/u];

%% Velocity definition
perturbationCase = 3;
derivative = @(t,x,~)derivative(t,x,false,u,lengthX,lengthY,epsilon,perturbationCase);
incompressible = true;
periodicBc = [true,false];
\end{lstlisting}

@@

To calculate the Cauchy-Green eigenvalues and eigenvectors we therefore set the grid resolution to $500 \times 199$. We also specify the flow is incompressible. Commands to obtain lambda-lines are given in listing~\ref{l:Bickley jet lambda-line commands}. The primary input needed are the positions of the Poincare sections which we determine

\begin{lstlisting}[caption={LCS Tool commands Bickley jet lambda-lines, subset of \texttt{demo/bickley\_jet/hyperbolic\_shear\_lcs.m}.},label=l:Bickley jet lambda-line commands]
lambda = 1;
lambdaLineOdeSolverOptions = odeset('relTol',1e-4);

poincareSection = struct('endPosition',{},'numPoints',{},'orbitMaxLength',{});
poincareSection(1).endPosition = [6.5,-1.4;5.2,-1.4]*1e6;
poincareSection(2).endPosition = [1.35e7,-1.4e6;1.5e7,-.5e6];
poincareSection(3).endPosition = [3.25,1.5;1.4,2.6]*1e6;
poincareSection(4).endPosition = [1e7,1.5e6;8e6,2.6e6];
poincareSection(5).endPosition = [1.65e7,1.5e6;1.5e7,2.6e6];

[shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda);
closedLambdaLine = poincare_closed_orbit_multi(domain,resolution,shearline,poincareSection,'odeSolverOptions',lambdaLineOdeSolverOptions,'periodicBc',periodicBc);
\end{lstlisting}

To obtain strainline hyperbolic LCS, we must set a local maximisation distance and a maximum LCS length. To demonstrate the
